### 2018-05-16 submission/manuscript.tex
remove copious mention of HiFi DNA from manuscript

\textbf{Background.} Bias is introduced at different stages of the 16S
rRNA gene sequencing workflow. Commonly studied sources of bias include
preservation and DNA extraction methods. Although the number of cycles
and high fidelity (HiFi) DNA polymerase are studied less often, they are
still important sources of bias in this workflow. Here, we examine how
both cycle number and HiFi DNA polymerase can change the bacterial
community and introduce bias to the final obtained results.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. The
analysis initially examined the number of Operational Taxonomic Units
(OTUs) for both fecal samples and mock communities. It also assessed
HiFi DNA polymerase dependent differences in the Bray-Curtis index,
error rate, sequence error prevalence, and chimera prevalence. Our
analysis also examined chimera prevalence correlation with the number of
OTUs.

\textbf{Results.} When analyzing fecal samples we observed a different
number of OTUs between HiFi DNA polymerases at 35 cycles (P-value
\textless{} 0.0001). Our analysis identified these HiFi dependent
differences in the number of OTUs as early as 20 cycles in the mock
communities (P-value = 0.002). Chimera prevalence varied by HiFi DNA
polymerase and this variation persisted after chimera removal using
VSEARCH. We also observed positive correlations between chimera
prevalence and the number of OTUs which was not affected by chimera
removal with VSEARCH.

\textbf{Conclusions.} HiFi DNA polymerase dependent differences in the
number of OTUs and chimera prevalence makes comparison across studies
difficult. Care should be exercised when choosing both HiFi DNA
polymerase and number of cycles to be used in 16S rRNA gene sequencing
studies.

@@

A recent study found clear differences between normal and high fidelity
(HiFi) DNA polymerase and that optimization of the PCR protocol could
reduce error and chimera generation (Gohl et al., 2016). This study also
found that regardless of DNA polymerase, the number of Operational
Taxonomic Units (OTUs) or taxa generated were not easily reduced using
the authors chosen bioinformatic pipeline (Gohl et al., 2016). It is
natural to extend this line of inquiry and ask if biases in the number
of OTUs and chimeras are also dependent on the type of HiFi DNA
polymerase. There is some reason to think that this may be the case
since many of these HiFi DNA polymerases come from different families
(e.g. \emph{Taq} belongs to the family A polymerases) and may
intrinsically have different error rates that cannot be completely
removed with modifications (Ishino \& Ishino, 2014).

Although bias introduced due to differences between DNA polymerase and
HiFi DNA polymerase has been investigated for 16S rRNA gene sequencing,
the bias caused due to differences between specific HiFi DNA polymerases
has not been. This study will specifically address how HiFi DNA
polymerases can bias observed bacterial community results derived from
16S rRNA gene sequencing. We will accomplish this by examining if any of
five different types of HiFi DNA polymerases introduce significant
biases into 16S rRNA gene surveys, if this is a cycle dependent
phenomenon, and whether they can be removed using a standard
bioinformatic pipeline.

@@

\textbf{\emph{PCR Protocol:}} The five different HiFi DNA polymerases
that were tested included AccuPrime\textsuperscript{TM} (ThermoFisher,
MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA),
Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA).
The PCR cycle conditions for Platinum and Accuprime followed a
previously published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
The HiFi DNA polymerase activation time was 2 minutes, unless a
different activation was specified. For Kappa and Q5, a previously
published protocol was used (Gohl et al., 2016). For Phusion, the
company defined conditions were used except for extension time, where
the Accuprime and Platinum settings were used.

The cycle conditions for both fecal and mock samples started at 15 and
increased by 5 up to 35 cycles with amplicons used at each 5-step
increase for sequencing. The PCR of fecal DNA samples consisted of all 4
samples at 15, 20, 25, 30, and 35 cycles for each HiFi DNA polymerase
(total sample n=100). The mock communities had 4 replicates at 15, 20,
25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA
polymerases (total samples n=130). No mock community sample had enough
PCR product at 15 cycles for adequate 16S rRNA gene sequencing.

@@

\textbf{\emph{Analysis Workflow:}} All samples were rarefied to 1000
sequence. The total number of OTUs was analyzed for both the fecal and
mock community samples. For fecal samples, cycle dependent affects on
Bray-Curtis indices were assessed for cycle group and within individual
differences from the previous cycle (e.g.~20 versus 25, 25 versus 30).
Based on these observations we analyzed potential reasons for these
differences. Analysis of the mock community of each HiFi DNA polymerase
for general sequence error rate, number of sequences with an error, base
substitution, and numbers of chimeras were assessed before and after
chimera removal. Additionally, the correlation between the number of
chimeras and the number of OTUs was also assessed.

@@

\textbf{\emph{The Number of OTUs are Dependent on HiFi DNA Polymerase:}}
A consistent difference in the range of number of OTUs, that was
dependent on the HiFi DNA polymerase used was observed for fecal samples
{[}Figure 1{]}. Additionally, there was a trend for lower number of
cycles (15-20) to result in a reduced range in the number of OTUs versus
higher number of cycles (25, 30, and 35) between HiFi DNA polymerases
{[}Figure 1{]}. For fecal samples, there was a significant difference
between HiFi DNA polymerases at 35 cycles (F-stat \textgreater{} 16.35,
P-value = 9.7e-05) {[}Table S1{]}. No differences between the specific
groups was found based on a Tukey post-hoc test (P-value \textgreater{}
0.05) {[}Table S2{]}. This HiFi DNA polymerase dependent difference in
the range of number of OTUs also was observed in the mock community
samples {[}Figure 2{]}. Regardless if fecal or mock communities were
used, the same HiFi DNA polymerases had high (Platinum) and low
(Accuprime) number of OTUs and this was consistent across the number of
cycles used {[}Figure 1-2 \& Table S1-S4{]}. In contrast to the results
obtained with fecal samples, differences between HiFi DNA polymerases
were observed as early as 20 cycles in the mock community samples
(F-stat = 15.82, P-value = 0.002) {[}Table S3{]}. For the mock
community, the majority of differences in the number of OTUs were
between Kappa and Platinum and the other HiFi DNA polymerases across
different number of cycles {[}Table S4{]}. Based on these observations
in fecal and mock communities, it is clear that different HiFi DNA
polymerases result in a different total number of OTUs within a sample.

\textbf{\emph{Minimal Bray-Curtis Differences are Detected and are
Dependent on Cycle Number:}} A few small differences based on the number
of cycles were detected in overall bacterial community composition.
Within the same fecal sample and independent of HiFi DNA polymerases,
there were large differences in the community composition between 20 and
25 cycles {[}Figure 3A{]}. Additionally, when data was available for the
mock communities, there were also large observed differences between 20
and 25 cycles {[}Figure 3B{]}. These community differences within a
sample disappear when comparing 25 to 30 cycles and do not persist past
25 cycles {[}Figure 3{]}. Although these trends are clearly noticeable,
we found that there was no detectable difference in Bray-Curtis index
when comparing to the previous 5-cycle increment for both fecal and mock
communities after multiple comparison correction (P-value \textgreater{}
0.05). Using PERMANOVA to test for community differences based on the
number of cycles within HiFi DNA polymerases, only Phusion had cycle
dependent differences (P-value = 0.03. For fecal samples, Phusion was
one of two HiFi DNA polymerases that had enough sequences to be rarefied
to 1000 at 15 cycles. Overall, these data suggest that there are small
HiFi DNA polymerase differences in Bray-Curtis index that are dependent
on the number of cycles.

\textbf{\emph{Sequence Error is Dependent on both HiFi DNA Polymerase
and Cycle Number:}} Differences in the median error rate varied by HiFi
DNA polymerase across the number of cycles used {[}Table S5{]}. The
highest error rates were for the Kappa HiFi DNA polymerase {[}Figure
4{]}. The differences in the median error rate between the different
HiFi DNA polymerases was dependent on the number of cycles with Platinum
and Kappa having the most differences versus other HiFi DNA polymerases
{[}Figure 4 and Table S6{]}. The total sequences with at least one error
was also cycle number dependent {[}Table S7 \& S8. These differences in
sequences with at least one error were mostly due to differences in
Accuprime\textsuperscript{TM} and Kappa versus the other HiFi DNA
polymerases {[}Table S7 \& S8{]}. Finally, we did not observe a HiFi DNA
polymerase dependent difference on base substitution rate {[}Figure
S1{]}. These results suggest that sequence error is dependent on HiFi
DNA polymerase. We next wanted to assess whether chimeras also were
dependent on HiFi DNA polymerase and whether this affected OTU number.

\textbf{\emph{Prevalence of Chimeric Sequences are HiFi DNA Polymerase
Dependent and Correlate with the Number of OTUs:}} There were
significant differences in the chimera prevalence based on HiFi DNA
polymerase used at all numbers of cycles used (P-value \textless{} 0.05)
{[}Table S9{]}. Differences in chimera prevalence between Platinum and
all other HiFi DNA polymerases accounted for the majority of these
differences {[}Table S10{]}. Accuprime\textsuperscript{TM} had the
lowest chimera prevalence of all HiFi DNA polymerases regardless of
whether chimera removal with VSEARCH was used {[}Figure 5A \& 5B{]}.
There was a plateau in the total percent of chimeras that were removed
that was similar for all HiFi DNA polymerases {[}Figure 5C{]}. A
positive correlation was observed between chimeric sequences and the
number of OTUs for all HiFi DNA polymerases {[}Figure 6{]}. This
positive correlation was strongest for Accuprime\textsuperscript{TM},
Platinum, and Phusion HiFi DNA Polymerases {[}Figure 6{]}. This data
suggests that chimera prevalence depends on HiFi DNA polymerase used and
confirms that the number of OTUs is dependent on the prevalence of these
chimeric sequences.

@@

In this study we show that the number of OTUs, error rate, and chimera
prevalence depends on HiFi DNA polymerase {[}Figure 1-2 \& 4-5{]}. These
differences are important because many diversity metrics rely on the
number of OTUs or other measures dependent on error rate and chimera
prevalence as part of their metric calculations (e.g.~richness).
Additionally, the earlier detection of differences in total number of
OTUs between HiFi DNA polymerases in the mock versus fecal samples might
indicate that high biomass samples may underestimate the biases present
within low biomass samples. We observed that undetected chimeras that
were not identified and removed using standard bioinformatic approaches
cause many of these differences. This suggests that specific diversity
differences between studies can be attributed to differences in HiFi DNA
polyermase used. Based on our observations metrics that measure within
sample diversity depend on HiFi DNA polymerase but this may not be the
case for metrics that assess between sample diversity.

There were few differences that depend on HiFi DNA polymerase for
between sample diversity, as measured by the Bray-Curtis index. Using
this metric our observations generally found no differences in the
overall bacterial community composition for the number of cycles used.
One possible reason for this outcome was that our study did not have
enough power to detect differences due to low sample number in each
group. Another reason was that many of the OTUs are likely not highly
abundant, allowing the Bray-Curtis index to be able to successfully
down-weight chimeric OTUs (Minchin, 1987). The choice of downstream
diversity metric could be an important consideration in helping to
mitigate these observed HiFi DNA polymerase dependent differences in
chimera prevalence. Metrics that solely use presence/absence of OTUs
(e.g.~Jaccard (Real \& Vargas, 1996)) may be less robust to chimera
prevalence and by extension total number of OTU differences in HiFi DNA
Polymerases. When choosing a distance metric, careful consideration of
the biases introduced from the PCR step of the 16S rRNA gene sequencing
workflow need to be taken into account. With differences in the number
of OTUs and chimera prevalence depending on HiFi DNA polymerase used, it
might be easier to avoid specific DNA polymerase families altogether.

Although the variation in error rate and chimera prevalence may be due
to the DNA polymerase family, the highest and lowest chimera rates both
belonged to a family A polymerase (Platinum and
Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
Additionally, based on the material safety data sheet (MSDS) the
differences between the two HiFi DNA polymerases are not immediately
apparent. Both HiFi DNA polymerases contain a recombinant \emph{Taq} DNA
polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
\emph{Taq} antibody. With everything else being equal, it is possible
that differences in how the recombinant \emph{Taq} was generated could
be a contributing factor for the differences observed between the HiFi
DNA polymerases. We are unlikely to avoid adding HiFi dependent bias to
16S rRNA gene sequencing results, however, these differences may not be
large enough to mask the actual biological signal.

The majority of HiFi DNA polymerases we studied add small increases in
the number of OTUs and chimera prevalence that can be masked by
biological differences. The sequence error introduced by the HiFi DNA
polymerase is small and likely to be smaller than the biological
variation within a specific study, which would be consistent with
previous findings for preservation and DNA extraction methods (Salter et
al., 2014; Song et al., 2016; Luo et al., 2016). However, the chimarea
prevalence for some HiFi DNA polymerases (e.g.~Platinum) are relatively
large and might be greater than the oberseved biological variation
within a specific study. The choice of HiFi DNA polymerase can be as
important a consideration as either preservation or DNA extraction
method used because similar to using different preservation methods or
different DNA extraction kits, the type of HiFi DNA polymerase can add
bias to the observed bacterial community. Although avoiding HiFi DNA
polymerases that yield a high number of OTUs and chimera prevalence
might provide better standardization of results it does come at a cost.

Methods can be standardized but they commonly contain bias that is
reproducible and may miss important associations. Bias can be easily
reproduced and can be found in every step of the 16S rRNA gene
sequencing workflow (Salter et al., 2014; Gohl et al., 2016; Luo et al.,
2016; Amir et al., 2017). This study shows that specific diversity
metrics used to measure the microbial community consistently vary based
on HiFi DNA polymerase. Standardizing multiple workflows to one specific
HiFi DNA polymerase could be detrimental since some HiFi DNA polymerases
may work better in certain situations over others. Arguably, the degree
of workflow standardization across studies and research group needs to
be approached on a study by study basis and not every project needs to
use the exact same approach. All aspects of the 16S rRNA gene sequencing
workflow need to be customized for the specific microbial community that
will be sampled. Although a diversity of approaches may make
reproducibility and replicability more difficult it will help to avoid
systematic biases from occuring due to widespread standardization of
approaches.

>>>>

\textbf{Background.} Bias is introduced at different stages of the 16S
rRNA gene sequencing workflow. Commonly studied sources of bias include
preservation and DNA extraction methods. Although the number of cycles
and high fidelity (HiFi) DNA polymerase are studied less often, they are
still important sources of bias in this workflow. Here, we examine how
both cycle number and polymerase can change the bacterial community and
introduce bias to the final obtained results.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. The
analysis initially examined the number of Operational Taxonomic Units
(OTUs) for both fecal samples and mock communities. It also assessed
polymerase dependent differences in the Bray-Curtis index, error rate,
sequence error prevalence, and chimera prevalence. Our analysis also
examined chimera prevalence correlation with the number of OTUs.

\textbf{Results.} When analyzing fecal samples we observed a different
number of OTUs between HiFi DNA polymerases at 35 cycles (P-value
\textless{} 0.0001). Our analysis identified these polymerase dependent
differences in the number of OTUs as early as 20 cycles in the mock
communities (P-value = 0.002). Chimera prevalence varied by polymerase
and this variation persisted after chimera removal using VSEARCH. We
also observed positive correlations between chimera prevalence and the
number of OTUs which was not affected by chimera removal with VSEARCH.

\textbf{Conclusions.} HiFi DNA polymerase dependent differences in the
number of OTUs and chimera prevalence makes comparison across studies
difficult. Care should be exercised when choosing the polymerase and
number of cycles to be used in 16S rRNA gene sequencing studies.

@@

A recent study found clear differences between normal and high fidelity
DNA polymerase and that optimization of the PCR protocol could reduce
error and chimera generation (Gohl et al., 2016). This study also found
that regardless of DNA polymerase, the number of Operational Taxonomic
Units (OTUs) or taxa generated were not easily reduced using the authors
chosen bioinformatic pipeline (Gohl et al., 2016). It is natural to
extend this line of inquiry and ask if biases in the number of OTUs and
chimeras are also dependent on the type of high fidelity DNA polymerase.
There is some reason to think that this may be the case since many of
these HiFi DNA polymerases come from different families (e.g. \emph{Taq}
belongs to the family A polymerases) and may intrinsically have
different error rates that cannot be completely removed with
modifications (Ishino \& Ishino, 2014).

Although bias introduced due to differences between DNA polymerase and
high fidelity DNA polymerase has been investigated for 16S rRNA gene
sequencing, the bias caused due to differences between specific high
fidelity DNA polymerases has not been. This study will specifically
address how high fidelity DNA polymerases can bias observed bacterial
community results derived from 16S rRNA gene sequencing. We will
accomplish this by examining if any of five different types of high
fidelity DNA polymerases introduce significant biases into 16S rRNA gene
surveys, if this is a cycle dependent phenomenon, and whether they can
be removed using a standard bioinformatic pipeline.

@@

\textbf{\emph{PCR Protocol:}} The five different high fidelity DNA
polymerases (hereto referred to as polymerases) that were tested
included AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA), KAPA
HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum
(ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR
cycle conditions for Platinum and Accuprime followed a previously
published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
The polymerases activation time was 2 minutes, unless a different
activation was specified. For Kappa and Q5, a previously published
protocol was used (Gohl et al., 2016). For Phusion, the company defined
conditions were used except for extension time, where the Accuprime and
Platinum settings were used.

The cycle conditions for both fecal and mock samples started at 15 and
increased by 5 up to 35 cycles with amplicons used at each 5-step
increase for sequencing. The PCR of fecal DNA samples consisted of all 4
samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total
sample n=100). The mock communities had 4 replicates at 15, 20, 25, and
35 cycles and 10 replicates for 30 cycles for all polymerases (total
samples n=130). No mock community sample had enough PCR product at 15
cycles for adequate 16S rRNA gene sequencing.

@@

\textbf{\emph{Analysis Workflow:}} All samples were rarefied to 1000
sequence. The total number of OTUs was analyzed for both the fecal and
mock community samples. For fecal samples, cycle dependent affects on
Bray-Curtis indices were assessed for cycle group and within individual
differences from the previous cycle (e.g.~20 versus 25, 25 versus 30).
Based on these observations we analyzed potential reasons for these
differences. Analysis of the mock community of each polymerase for
sequence error rate, number of sequences with an error, base
substitution, and numbers of chimeras before and after chimera removal
with VSEARCH was assessed. Additionally, the correlation between the
number of chimeras and the number of OTUs was also assessed.

@@

\textbf{\emph{The Number of OTUs are Dependent on Polymerase:}} A
consistent difference in the range of number of OTUs, that was dependent
on the polymerase used was observed for fecal samples {[}Figure 1{]}.
Additionally, there was a trend for lower number of cycles (15-20) to
result in a reduced range in the number of OTUs versus higher number of
cycles (25, 30, and 35) between polymerases {[}Figure 1{]}. For fecal
samples, there was a significant difference between polymerases at 35
cycles (F-stat \textgreater{} 16.35, P-value = 9.7e-05) {[}Table S1{]}.
No differences between the specific groups was found based on a Tukey
post-hoc test (P-value \textgreater{} 0.05) {[}Table S2{]}. This
polymerase dependent difference in the range of number of OTUs also was
observed in the mock community samples {[}Figure 2{]}. Regardless if
fecal or mock communities were used, the same polymerases had high
(Platinum) and low (Accuprime) number of OTUs and this was consistent
across the number of cycles used {[}Figure 1-2 \& Table S1-S4{]}. In
contrast to the results obtained with fecal samples, differences between
polymerases were observed as early as 20 cycles in the mock community
samples (F-stat = 15.82, P-value = 0.002) {[}Table S3{]}. For the mock
community, the majority of differences in the number of OTUs were
between Kappa and Platinum and the other polymerases across different
number of cycles {[}Table S4{]}. Based on these observations in fecal
and mock communities, it is clear that different polymerases result in a
different total number of OTUs within a sample.

\textbf{\emph{Minimal Bray-Curtis Differences are Detected and are
Dependent on Cycle Number:}} A few small differences based on the number
of cycles were detected in overall bacterial community composition.
Within the same fecal sample and independent of polymerases, there were
large differences in the community composition between 20 and 25 cycles
{[}Figure 3A{]}. Additionally, when data was available for the mock
communities, there were also large observed differences between 20 and
25 cycles {[}Figure 3B{]}. These community differences within a sample
disappear when comparing 25 to 30 cycles and do not persist past 25
cycles {[}Figure 3{]}. Although these trends are clearly noticeable, we
found that there was no detectable difference in Bray-Curtis index when
comparing to the previous 5-cycle increment for both fecal and mock
communities after multiple comparison correction (P-value \textgreater{}
0.05). Using PERMANOVA to test for community differences based on the
number of cycles within polymerases, only Phusion had cycle dependent
differences (P-value = 0.03. For fecal samples, Phusion was one of two
polymerases that had enough sequences to be rarefied to 1000 at 15
cycles. Overall, these data suggest that there are small polymerase
differences in Bray-Curtis index that are dependent on the number of
cycles.

\textbf{\emph{Sequence Error is Dependent on both Polymerase and Cycle
Number:}} Differences in the median error rate varied by polymerase
across the number of cycles used {[}Table S5{]}. The highest error rates
were for the Kappa HiFi DNA polymerase {[}Figure 4{]}. The differences
in the median error rate between the different polymerases was dependent
on the number of cycles with Platinum and Kappa having the most
differences versus other polymerases {[}Figure 4 and Table S6{]}. The
total sequences with at least one error was also cycle number dependent
{[}Table S7 \& S8. These differences in sequences with at least one
error were mostly due to differences in Accuprime\textsuperscript{TM}
and Kappa versus the other polymerases {[}Table S7 \& S8{]}. Finally, we
did not observe a polymerase dependent difference on base substitution
rate {[}Figure S1{]}. These results suggest that sequence error is
dependent on polymerase. We next wanted to assess whether chimeras also
were dependent on polymerase and whether this affected OTU number.

\textbf{\emph{Prevalence of Chimeric Sequences are Polymerase Dependent
and Correlate with the Number of OTUs:}} There were significant
differences in the chimera prevalence based on polymerase used at all
numbers of cycles used (P-value \textless{} 0.05) {[}Table S9{]}.
Differences in chimera prevalence between Platinum and all other
polymerases accounted for the majority of these differences {[}Table
S10{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
of all polymerases regardless of whether chimera removal with VSEARCH
was used {[}Figure 5A \& 5B{]}. There was a plateau in the total percent
of chimeras that were removed that was similar for all polymerases
{[}Figure 5C{]}. A positive correlation was observed between chimeric
sequences and the number of OTUs for all polymerases {[}Figure 6{]}.
This positive correlation was strongest for
Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 6{]}.
This data suggests that chimera prevalence depends on polymerase used
and confirms that the number of OTUs is dependent on the prevalence of
these chimeric sequences.

@@

In this study we show that the number of OTUs, error rate, and chimera
prevalence depends on polymerase {[}Figure 1-2 \& 4-5{]}. These
differences are important because many diversity metrics rely on the
number of OTUs or other measures dependent on error rate and chimera
prevalence as part of their metric calculations (e.g.~richness).
Additionally, the earlier detection of differences in total number of
OTUs between polymerases in the mock versus fecal samples might indicate
that high biomass samples may underestimate the biases present within
low biomass samples. We observed that undetected chimeras that were not
identified and removed using standard bioinformatic approaches cause
many of these differences. This suggests that specific diversity
differences between studies can be attributed to differences in
polyermase used. Based on our observations metrics that measure within
sample diversity depend on polymerase but this may not be the case for
metrics that assess between sample diversity.

There were few differences that depend on polymerase for between sample
diversity, as measured by the Bray-Curtis index. Using this metric our
observations generally found no differences in the overall bacterial
community composition for the number of cycles used. One possible reason
for this outcome was that our study did not have enough power to detect
differences due to low sample number in each group. Another reason was
that many of the OTUs are likely not highly abundant, allowing the
Bray-Curtis index to be able to successfully down-weight chimeric OTUs
(Minchin, 1987). The choice of downstream diversity metric could be an
important consideration in helping to mitigate these observed polymerase
dependent differences in chimera prevalence. Metrics that solely use
presence/absence of OTUs (e.g.~Jaccard (Real \& Vargas, 1996)) may be
less robust to chimera prevalence and by extension total number of OTU
differences in polymerases. When choosing a distance metric, careful
consideration of the biases introduced from the PCR step of the 16S rRNA
gene sequencing workflow need to be taken into account. With differences
in the number of OTUs and chimera prevalence depending on polymerase
used, it might be easier to avoid specific DNA polymerase families
altogether.

Although the variation in error rate and chimera prevalence may be due
to the DNA polymerase family, this is unlikely to be the only
contributor. For example, the highest and lowest chimera rates both
belonged to a family A polymerase (Platinum and
Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
Additionally, based on the material safety data sheet (MSDS) the
differences between the two polymerases are not immediately apparent.
Both polymerases contain a recombinant \emph{Taq} DNA polymerase, a
\emph{Pyrococcus} spp GB-D polymerase and a platinum \emph{Taq}
antibody. With everything else being equal, it is possible that
differences in how the recombinant \emph{Taq} was generated could be a
contributing factor for the differences observed between the
polymerases. We are unlikely to avoid adding polymerase dependent bias
to 16S rRNA gene sequencing results, however, these differences may not
be large enough to mask the actual biological signal.

The majority of polymerases we studied add small increases in the number
of OTUs and chimera prevalence that can be masked by biological
differences. The sequence error introduced by the polymerase is small
and likely to be smaller than the biological variation within a specific
study, which would be consistent with previous findings for preservation
and DNA extraction methods (Salter et al., 2014; Song et al., 2016; Luo
et al., 2016). However, the chimarea prevalence for some polymerases
(e.g.~Platinum) are relatively large and might be greater than the
oberseved biological variation within a specific study. The choice of
polymerase can be as important a consideration as either preservation or
DNA extraction method used because similar to using different
preservation methods or different DNA extraction kits, the type of
polymerase can add bias to the observed bacterial community. Although
avoiding polymerases that yield a high number of OTUs and chimera
prevalence might provide better standardization of results it does come
at a cost.

Methods can be standardized but they commonly contain bias that is
reproducible and may miss important associations. Bias can be easily
reproduced and can be found in every step of the 16S rRNA gene
sequencing workflow (Salter et al., 2014; Gohl et al., 2016; Luo et al.,
2016; Amir et al., 2017). This study shows that specific diversity
metrics used to measure the microbial community consistently vary based
on polymerase. Standardizing multiple workflows to one specific
polymerase could be detrimental since some polymerases may work better
in certain situations over others. Arguably, the degree of workflow
standardization across studies and research group needs to be approached
on a study by study basis and not every project needs to use the exact
same approach. All aspects of the 16S rRNA gene sequencing workflow need
to be customized for the specific microbial community that will be
sampled. Although a diversity of approaches may make reproducibility and
replicability more difficult it will help to avoid systematic biases
from occuring due to widespread standardization of approaches.
